---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---
### Neal Koss
### Jan 18, 2015

## Loading and preprocessing the data
```{r}
inp <- read.csv("activity.csv")
# change the 'date' column to a date format internally
inp$date <- as.Date(inp$date)

```

## What is mean total number of steps taken per day?
```{r}
# divide the 'steps' into groups based on the 'date'
#  then get the sum of the steps for each date
totalStepsPerDay <- sapply(split(inp$steps, inp$date), sum, na.rm=T)
meanPerDay <- mean(totalStepsPerDay)
medianPerDay <- median(totalStepsPerDay)

```
The mean total number of steps per day is `r meanPerDay` and median is `r medianPerDay`  
```{r}
hist(totalStepsPerDay, col="blue")

```

## What is the average daily activity pattern?
```{r}
# For each 5-minute interval (1:288), find the mean number of steps
meanStepsPerInterval <- aggregate(steps ~ interval, data=inp, FUN=mean, na.action=na.omit)
intervals<- unique(inp$interval)
plot(x=intervals, y=meanStepsPerInterval$steps, type='l', col="blue", xlab='5-minute intervals', ylab='mean number of steps')

# find the 5-minute interval with the maximum number of steps
#  and how many steps in that interval
maxInterval <- which.max(meanStepsPerInterval$steps)
maxSteps <- meanStepsPerInterval$steps[maxInterval]

```
The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is `r maxInterval` and it contains `r maxSteps` steps

## Imputing missing values
```{r}
# Find number of rows with missing data
missing <- nrow(inp) - nrow(na.omit(inp))

```
There are `r missing` rows with missing data in the dataset

```{r}
# Fill in missing values (steps) with the mean steps per interval
newInp <- inp
for (i in 1:nrow(inp)) {
    if (is.na(inp$steps[i]))
        newInp$steps[i] <- meanStepsPerInterval$steps[(i %% 288) + 1]
}

# Find the 'new' total steps per day, plot and get summaries
newTotalStepsPerDay <- sapply(split(newInp$steps, newInp$date), sum)
hist(newTotalStepsPerDay, col="blue")

newMeanPerDay <- mean(newTotalStepsPerDay)
newMedianPerDay <- median(newTotalStepsPerDay)

```
The new mean number of steps per day is `r newMeanPerDay`
The new median is `r newMedianPerDay`  

These values are definitely different from the values using the dataset when omitting missing data. Both values are greater and the data is closer to a normal distribution since the mean and median are equal.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
# Function to label weekend or weekday for each date
daytype <- function(date) {
    if (weekdays(as.Date(date)) %in% c("Saturday", "Sunday")) {
        "weekend"
    } else {
        "weekday"
    }
}

# Add the dayType as an extra column
newInp$dayType <- as.factor(sapply(newInp$date, daytype))

# Plot the weekend and weekdayinterval means
par(mfrow=c(2,1))
for (type in c("weekend", "weekday")) {
    steps.type <- aggregate(steps ~ interval, data=newInp, 
                            subset=newInp$dayType==type, FUN=mean)
    plot(steps.type, type="l", main=type)
}

```

Note that on weekends the activity tends to be more spread out over time, presumably since people are available for exercise at all times since it is not a workday.
